Cloudbreak: A MapReduce Algorithm for Detecting Genomic Structural Variation

### 2013-06-02 manuscript/manuscript.tex

\begin{table}
\begin{center}
\begin{tabular}{r|rrr|rrrrr}
  \cline{2-9}
   &                     & Prec. & Recall & 40-100bp  & 101-250bp  & 251-500bp & 501-1000bp & $>$ 1000bp \\ 
\hline
\multirow{7}{*}{\begin{sideways}Deletions\end{sideways}} & Total Number &          &           & 224 &  84 & 82 &  31 & 26\\ 
  \hline
\cline{2-9}
&  Cloudbreak    &  0.532 & \textbf{0.66} & \textbf{146} (  6)  &  58 (\textbf{  4}) &   62 (  1) &  14 (  1) &  15 (0) \\ 
&  Breakdancer   &  0.356 & 0.49 &  89 (0)  &  54 (0) &   53 (0) &   8 (0) &  15 (0) \\ 
&  GASVPro        & 0.146 & 0.432 &  83 (  2)  &  32 (0) &   55 (0) &   8 (0) &  15 (0) \\ 
&  DELLY-RP           & 0.457 & 0.613 & 114 (  8)  & \textbf{ 68} (  2) &  \textbf{ 66} (0) &   9 (0) &  17 (0) \\ 
&  DELLY-SR           & \textbf{0.679} & 0.166 & 0 (0)  &   3 (0) &   49 (0) &   6 (0) &  16 (0) \\ 
&  Pindel           & 0.462 & 0.421 &  96 (\textbf{ 13})  &  24 (0) &   48 (0) &   5 (0) &  15 (0)\\ 
&  MoDIL           & 0.132  & \textbf{0.66} & 123 (  5)  &  66 (  2) &  \textbf{ 66} (\textbf{  7}) & \textbf{ 17} (\textbf{  6}) & \textbf{ 23} (\textbf{  8})\\ 
   \hline
\multirow{5}{*}{\begin{sideways}Insertions\end{sideways}} & Total Number &          &           & 199 &  83 & 79 &  21 & 21\\ 
\cline{2-9}
&  Cloudbreak   &\textbf{0.705} & \textbf{0.196}  & \textbf{ 67} ( 24)  &   8 (  3) &    3 (  1) &   1 (0) & 0 (0) \\ 
&  Breakdancer & 0.262 & 0.0968  &  23 (  8)  & \textbf{ 14} (\textbf{  9}) &    2 (  1) & 0 (0) & 0 (0) \\ 
&  Pindel          & 0.572 & \textbf{0.196} &  52 (\textbf{ 26})  &   5 (  3) &  \textbf{ 10} (\textbf{ 10}) & \textbf{  3} (\textbf{  3}) & \textbf{  9} (\textbf{  9})\\ 
&  MoDIL          & 0.186 & 0.0521 &  14 (  1)  &   4 (  1) &    1 (0) &   2 (  1) & 0 (0)\\ 
\hline
\end{tabular}
\end{center}
\caption{The precision and recall of each tool for the simulated deletions and insertions in the 30X diploid chromosome 2 with Venter indels at maximum sensitivity, as well as the number of those variants discovered in each size class. Variants exclusively predicted by one tool are shown in parentheses. The total number of variants in each size class in the true set of deletions and insertions is shown in the first row of each section.}
\label{chr2DeletionAndInsertionPredsMaxSensitivity}
\end{table}

@@

\begin{table}
\begin{center}
\begin{tabular}{r|rrr|rrrrr}
  \cline{2-9}
&  & Prec. & Recall & 40-100bp & 101-250bp & 251-500bp & 501-1000bp & $>$ 1000bp \\ 
\hline
\multirow{6}{*}{\begin{sideways}Deletions\end{sideways}} & Total Number & & & 7,466 & 235 & 218 & 110 & 375 \\
  \hline
\cline{2-9}
& Cloudbreak & 0.11 & \textbf{0.13} & \textbf{371} (\textbf{127})  & \textbf{146} (\textbf{ 15}) &  \textbf{177} (\textbf{  9}) & \textbf{ 95} (\textbf{  3}) & \textbf{301} (\textbf{ 17}) \\ 
& Breakdancer & 0.122 & 0.112 & 261 ( 41)  & 132 (  6) &  167 (0) &  92 (0) & 288 (  2) \\  
&  GASVPro & 0.135 & 0.0449 & 120 ( 28)  &  38 (  3) &   80 (0) &  29 (0) & 110 (0) \\ 
&  DELLY-RP & 0.0824 & 0.091 & 143 (  8)  & 125 (  4) &  158 (  1) &  83 (  1) & 256 (  1) \\ 
&  DELLY-SR & 0.135 & 0.0439 & 0 (0)  &  25 (0) &  117 (0) &  56 (0) & 171 (0) \\ 
&  Pindel & \textbf{0.16} & 0.0685 & 149 ( 12)  &  57 (0) &  140 (0) &  58 (0) & 172 (0) \\ 
\hline
\multirow{4}{*}{\begin{sideways}Insertions\end{sideways}} & Total Number & & & 536 & 114 & 45 & 1 \\
\cline{2-9}
& Cloudbreak & 0.0268 & \textbf{0.391} & \textbf{246} (\textbf{ 83})  &  25 (  9) &    1 (0) & 0 (0) \\ 
& Breakdancer & 0.0281 & 0.181 &  97 ( 11)  & \textbf{ 27} (  9) &    2 (  1) & 0 (0) \\  
&  Pindel & \textbf{0.0387} & 0.239 & 144 ( 41)  &  14 (\textbf{ 12}) &  \textbf{  7} (\textbf{  7}) & \textbf{  1} (\textbf{  1})  \\ 
\hline
\end{tabular}
\end{center}
\caption{The precision and recall with respect to the gold standard set of deletions and insertions for each tool on the NA18507 data, as well as the number of variants found in each size class found. Exclusive predictions are in parentheses. For deletions, the same cutoffs were used as for the simulated data as in Table \ref{chr2DeletionPredsFDR10}, adjusted for the difference in coverage from 30X to 37X. For insertions, the maximum sensitivity cutoff was used.}
\label{NA18507DeletionAndInsertionPreds}
\end{table}

@@

@@

\begin{table}
\begin{center}
\begin{tabular}{rrr|rr}
 & \multicolumn{2}{c}{Simulated Data} & \multicolumn{2}{c}{NA18507} \\
\hline
 &  Non-repeat & Repeat  &  Non-repeat & Repeat \\ 
 Total Number & 120 & 327 & 553 & 7851 \\ 
  \hline
  Cloudbreak  &  25 (  2) & \textbf{158} ( 17) & \textbf{220} (\textbf{ 49}) & \textbf{870} (\textbf{124}) \\ 
  Breakdancer & \textbf{ 29} (  6) & 142 (  7) & 186 ( 19) & 754 ( 30) \\
  GASVPro     &  16 (  2) &  79 (  2) &  75 (  9) & 302 ( 22) \\
  DELLY       &  21 (  2) & 117 (  2) & 147 (  9) & 618 (  8) \\
  Pindel      &  18 (\textbf{  9}) & 112 (\textbf{ 26}) & 103 (  4) & 473 (  8) \\ 
   \hline
\end{tabular}
\end{center}
\caption{Detected deletions on the simulated and NA18507 data sets identified by each tool, broken down by whether the deletion overlaps with a RepeatMasker-annotated element.}
\label{deletionRepmaskpreds}
\end{table}

\begin{table}
\begin{center}
\begin{tabular}{rrr|rr}
 & \multicolumn{2}{c}{Simulated Data} & \multicolumn{2}{c}{NA18507} \\
\hline
 &  Non-repeat & Repeat  &  Non-repeat & Repeat \\ 
 Total Number & 133 & 270 & 341 & 355 \\ 
  \hline
  Cloudbreak  &  21 (  4) & \textbf{ 58} ( 24) & \textbf{137} (\textbf{ 37}) & \textbf{135} (\textbf{ 55}) \\ 
  Breakdancer &  17 (  7) &  22 ( 11) &  82 ( 12) &  44 (  9) \\
  Pindel      & \textbf{ 25} (\textbf{ 18}) &  54 (\textbf{ 33}) &  84 ( 36) &  82 ( 25) \\ 
  MoDIL      &   5 (0) &  16 (  3) & NA & NA \\ 
   \hline
\end{tabular}
\end{center}
\caption{Detected insertions in the simulated and NA18507 data sets identified by each tool, broken down by whether the insertion occurs in a RepeatMasker-annotated element.}
\label{insertionRepmaskpreds}
\end{table}

@@

Since Cloudbreak explicitly fits a mixture of insert size distributions at each location, and each component of the mixture represents an allele containing or not containing an insertion or deletion, it should be possible to use the parameters of the fit GMM to infer the genotype of each predicted variant, assuming that our pipeline is capturing all relevant read mappings near the locus of the variant. To test this, we attempted to classify the true positive Cloudbreak predictions described above for both the simulated and biological data sets. Because we simulated a diploid genome, we had genotypes for the simulated data readily available. For benchmarking genotyping of deletions in the real dataset, we only considered the deletions from the 1000 Genomes Project data set, which had been genotyped as part of the Phase 1 release using the population-scale SV detection algorithm Genome STRiP \autocite{Handsaker:2011ki}. We found that we could use $\alpha$, the mixing parameter that controls the weight of the two components in the GMM, to accurately predict deletion genotypes. By setting a simple cutoff of .35 on the average value of $\alpha$ in each Cloudbreak prediction, we were able to achieve 88.1\% and 96.5\% accuracy in predicting the genotype of the true positive deletions we detected at our 10\% FDR threshold in the simulated and real data sets respectively. Table \ref{deletionGenotypeaccuracy} shows confusion matrices for the two samples using this classifier. For insertions, none of the three input sets that made up the gold standard for NA18507 contained a sufficient number of insertion rates that meet our size threshold that also had genotyping information. Of the 79 insertions detected by Cloudbreak on the simulated data set, 13 were heterozygous. Using the same cutoff of .35 on $\alpha$, Cloudbreak correctly classified 62 of the 66 homozygous insertions and 10 of the 13 heterozygous insertions discovered, for an overall accuracy rate of 91\%.

\begin{table}
\begin{center}
\begin{tabular}{r|r|rr|rr|}
\multicolumn{2}{c}{}  & \multicolumn{4}{c}{Actual Genotypes} \\
\multicolumn{2}{c}{}  & \multicolumn{2}{c}{Simulated Data} & \multicolumn{2}{c}{NA18507} \\
\cline{3-6}
\multicolumn{2}{c|}{} &  Homozygous & Heterozygous & Homozygous & Heterozygous \\ 
\cline{2-6}
\multirow{2}{*}{\shortstack{Predicted \\ Genotypes}} & Homozygous & 100 & 4 &  77 & 12 \\
 & Heterozygous & 7 & 72 &  3 & 338 \\
\cline{2-6}
\end{tabular}
\end{center}
\caption{Confusion matrices for the predicted genotype of deletions found by Cloudbreak on both the simulated and NA18507 data sets.}
\label{deletionGenotypeaccuracy}
\end{table}

>>>>

\begin{table}
\begin{center}
\begin{tabular}{r|rrr|rrrrr}
  \cline{2-9}
   &                     & Prec. & Recall & 40-100bp  & 101-250bp  & 251-500bp & 501-1000bp & $>$ 1000bp \\ 
\hline
\multirow{7}{*}{\begin{sideways}Deletions\end{sideways}} & Total Number &          &           & 224 &  84 & 82 &  31 & 26\\ 
  \hline
\cline{2-9}
&  Cloudbreak    &  0.532 & \textbf{0.66} & \textbf{146} (6)  & 58 (\textbf{4}) &  62 (1) & 14 (1) & 15 (0) \\ 
&  Breakdancer   &  0.356 & 0.49 & 89 (0)  & 54 (0) &  53 (0) & 8 (0) & 15 (0) \\ 
&  GASVPro        & 0.146 & 0.432 & 83 (2)  & 32 (0) &  55 (0) & 8 (0) & 15 (0) \\ 
&  DELLY-RP           & 0.457 & 0.613 & 114 (8)  & \textbf{68} (2) &  \textbf{66} (0) & 9 (0) & 17 (0) \\ 
&  DELLY-SR           & \textbf{0.679} & 0.166 & 0 (0)  & 3 (0) &  49 (0) & 6 (0) & 16 (0) \\ 
&  Pindel           & 0.462 & 0.421 & 96 (\textbf{13})  & 24 (0) &  48 (0) & 5 (0) & 15 (0)\\ 
&  MoDIL           & 0.132  & \textbf{0.66} & 123 (5)  & 66 (2) &  \textbf{66} (\textbf{7}) & \textbf{17} (\textbf{6}) & \textbf{23} (\textbf{8})\\ 
   \hline
\multirow{5}{*}{\begin{sideways}Insertions\end{sideways}} & Total Number &          &           & 199 &  83 & 79 &  21 & 21\\ 
\cline{2-9}
&  Cloudbreak   &\textbf{0.705} & \textbf{0.196}  & \textbf{67} (24)  & 8 (3) &  3 (1) & 1 (0) & 0 (0) \\ 
&  Breakdancer & 0.262 & 0.0968  & 23 (8)  & \textbf{14} (\textbf{9}) &  2 (1) & 0 (0) & 0 (0) \\ 
&  Pindel          & 0.572 & \textbf{0.196} & 52 (\textbf{26})  & 5 (3) &  \textbf{10} (\textbf{10}) & \textbf{3} (\textbf{3}) & \textbf{9} (\textbf{9})\\ 
&  MoDIL          & 0.186 & 0.0521 & 14 (1)  & 4 (1) &  1 (0) & 2 (1) & 0 (0)\\ 
\hline
\end{tabular}
\end{center}
\caption{The number of simulated deletions and insertions in the 30X diploid chromosome 2 with Venter indels found by each tool at maximum sensitivity, as well as the number of those variants that were discovered exclusively by each tool (in parentheses). The total number of variants in each size class in the true set of deletions and insertions is shown in the first row of each section.}
\label{chr2DeletionAndInsertionPredsMaxSensitivity}
\end{table}

@@

\begin{table}
\begin{center}
\begin{tabular}{r|rrr|rrrrr}
  \cline{2-9}
&  & Prec. & Recall & 40-100bp & 101-250bp & 251-500bp & 501-1000bp & $>$ 1000bp \\ 
\hline
\multirow{6}{*}{\begin{sideways}Deletions\end{sideways}} & Total Number & & & 7,462 & 240 & 232 & 147 & 540 \\
  \hline
\cline{2-9}
& Cloudbreak & 0.124 & \textbf{0.142} & \textbf{371} (\textbf{127})  & \textbf{150} (\textbf{15}) &  \textbf{187} (\textbf{9}) & \textbf{123} (\textbf{8}) & \textbf{396} (\textbf{27}) \\ 
& Breakdancer & 0.137 & 0.123 & 261 (41)  & 136 (6) &  178 (0) & 114 (0) & 371 (2) \\  
&  GASVPro & 0.147 & 0.0474 & 120 (28)  & 40 (3) &  85 (0) & 36 (0) & 128 (0) \\ 
&  DELLY-RP & 0.0931 & 0.1 & 143 (8)  & 128 (4) &  167 (1) & 103 (1) & 323 (1) \\ 
&  DELLY-SR & 0.153 & 0.0485 & 0 (0)  & 26 (0) &  123 (0) & 66 (0) & 203 (0) \\ 
&  Pindel & \textbf{0.179} & 0.0748 & 149 (12)  & 61 (0) &  149 (0) & 69 (0) & 217 (0) \\ 
\hline
\multirow{4}{*}{\begin{sideways}Insertions\end{sideways}} & Total Number & & & 536 & 114 & 45 & 1 & 0 \\
\cline{2-9}
& Cloudbreak & 0.0268 & \textbf{0.391} & \textbf{246} (\textbf{83})  & 25 (9) &  1 (0) & 0 (0)  & 0 (0)  \\ 
& Breakdancer & 0.0281 & 0.181 & 97 (11)  & \textbf{27} (9) &  2 (1) & 0 (0) & 0 (0) \\  
&  Pindel & \textbf{0.0387} & 0.239 & 144 (41)  & 14 (\textbf{12}) &  \textbf{7} (\textbf{7}) & \textbf{1} (\textbf{1}) &  0 (0) \\ 
\hline
\end{tabular}
\end{center}
\caption{The precision and recall with respect to the gold standard set of deletions and insertions for each tool on the NA18507 data, as well as the number of variants found in each size class found. Exclusive predictions are in parentheses. For deletions, the same cutoffs were used as for the simulated data as in Table \ref{chr2DeletionPredsFDR10}, adjusted for the difference in coverage from 30X to 37X. For insertions, the maximum sensitivity cutoff was used.}
\label{NA18507DeletionAndInsertionPreds}
\end{table}

@@

\subsection{Performance on a Low-Coverage Cancer Data Set}

We also tested Cloudbreak on a sequencing data set obtained from a patient with acute myeloid leukemia (AML). We sequenced CD14+ cells on an Illumina Genome Analyzer II, producing 128,819,200 76bp paired end reads with a mean insert size of 285bp and standard deviation of 50bp, yielding sequence coverage of 5X and physical coverage of 8X. Using a pipeline consisting of Novoalign, Breakdancer, and a set of custom scripts for filtering and annotating candidate SVs, we had previously identified a set of variants present in this sample and validated several using PCR, including 8 deletions. Cloudbreak was able to identify all 8 of the validated deletions, showing that it is still sensitive to variants even with lower coverage data sets with a greater variance of insert sizes. The variants identfied include deletions occurring in the genes CTDSPL/RBPS3, an AML tumor suppressor \autocite{Zheng:2012kk} and CAV2, a modulator of cancer progression \autocite{Lee:2011bz}. We are currently investigating these deletions to determine their functional impact on this patient.

@@

\begin{table}[b]
\begin{center}
\begin{tabular}{rrr|rr}
 & \multicolumn{2}{c}{Simulated Data} & \multicolumn{2}{c}{NA18507} \\
\hline
 &  Non-repeat & Repeat  &  Non-repeat & Repeat \\ 
 Total Number & 120 & 327 & 562 & 8059 \\ 
  \hline
  Cloudbreak  & 25 (2) & \textbf{158} (17) & \textbf{226} (\textbf{49}) & \textbf{1001} (\textbf{137}) \\ 
  Breakdancer & \textbf{29} (6) & 142 (7) & 192 (19) & 868 (30) \\
  GASVPro     & 16 (2) & 79 (2) & 79 (9) & 330 (22) \\
  DELLY-RP       & 21 (2) & 117 (2) & 152 (9) & 712 (6) \\
  DELLY-SR       & 0 (0) & 42 (0) & 27 (0) & 391 (0) \\
  Pindel      & 18 (\textbf{9}) & 112 (\textbf{26}) & 109 (4) & 536 (8) \\ 
   \hline
\end{tabular}
\end{center}
\caption{Detected deletions on the simulated and NA18507 data sets identified by each tool, broken down by whether the deletion overlaps with a RepeatMasker-annotated element.}
\label{deletionRepmaskpreds}
\end{table}

\begin{table}[b]
\begin{center}
\begin{tabular}{rrr|rr}
 & \multicolumn{2}{c}{Simulated Data} & \multicolumn{2}{c}{NA18507} \\
\hline
 &  Non-repeat & Repeat  &  Non-repeat & Repeat \\ 
 Total Number & 133 & 270 & 341 & 355 \\ 
  \hline
  Cloudbreak  & 21 (4) & \textbf{58} (24) & \textbf{137} (\textbf{37}) & \textbf{135} (\textbf{55}) \\ 
  Breakdancer & 17 (7) & 22 (11) & 82 (12) & 44 (9) \\
  Pindel      & \textbf{25} (\textbf{18}) & 54 (\textbf{33}) & 84 (36) & 82 (25) \\ 
  MoDIL      & 5 (0) & 16 (3) & NA & NA \\ 
   \hline
\end{tabular}
\end{center}
\caption{Detected insertions in the simulated and NA18507 data sets identified by each tool, broken down by whether the insertion occurs in a RepeatMasker-annotated element.}
\label{insertionRepmaskpreds}
\end{table}

@@

Since Cloudbreak explicitly fits a mixture of insert size distributions at each location, and each component of the mixture represents an allele containing or not containing an insertion or deletion, it should be possible to use the parameters of the fit GMM to infer the genotype of each predicted variant, assuming that our pipeline is capturing all relevant read mappings near the locus of the variant. To test this, we attempted to classify the true positive Cloudbreak predictions described above for both the simulated and biological data sets. Because we simulated a diploid genome, we had genotypes for the simulated data readily available. For benchmarking genotyping of deletions in the real dataset, we only considered the deletions from the 1000 Genomes Project data set, which had been genotyped as part of the Phase 1 release using the population-scale SV detection algorithm Genome STRiP \autocite{Handsaker:2011ki}. We found that we could use $\alpha$, the mixing parameter that controls the weight of the two components in the GMM, to accurately predict deletion genotypes. By setting a simple cutoff of .25 on the average value of $\alpha$ in each Cloudbreak prediction, we were able to achieve 86.8\% and 96.0\% accuracy in predicting the genotype of the true positive deletions we detected at our 10\% FDR threshold in the simulated and real data sets respectively. Table \ref{deletionGenotypeaccuracy} shows confusion matrices for the two samples using this classifier. For insertions, none of the three input sets that made up the gold standard for NA18507 contained a sufficient number of insertion rates that meet our size threshold that also had genotyping information. Of the 79 insertions detected by Cloudbreak on the simulated data set, 13 were heterozygous. Using the same cutoff of .35 on $\alpha$, Cloudbreak correctly classified 62 of the 66 homozygous insertions and 10 of the 13 heterozygous insertions discovered, for an overall accuracy rate of 91\%.

\begin{table}
\begin{center}
\begin{tabular}{r|r|rr|rr|}
\multicolumn{2}{c}{}  & \multicolumn{4}{c}{Actual Genotypes} \\
\multicolumn{2}{c}{}  & \multicolumn{2}{c}{Simulated Data} & \multicolumn{2}{c}{NA18507} \\
\cline{3-6}
\multicolumn{2}{c|}{} &  Homozygous & Heterozygous & Homozygous & Heterozygous \\ 
\cline{2-6}
\multirow{2}{*}{\shortstack{Predicted \\ Genotypes}} & Homozygous & 100 & 4 &  95 & 19 \\
 & Heterozygous & 7 & 72 &  4 & 451 \\
\cline{2-6}
\end{tabular}
\end{center}
\caption{Confusion matrices for the predicted genotype of deletions found by Cloudbreak on both the simulated and NA18507 data sets.}
\label{deletionGenotypeaccuracy}
\end{table}
